---
title: 'Did Nebraska deserve to have a 3-9 season? '
author: Maddi Anderson
date: '2022-04-08'
slug: did-nebraska-deserve-to-have-a-3-9-season
categories: []
tags: []
---
```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(cfbfastR)
library(scales)
library(ggrepel)
library(ggalt)
```
Nebraska fans are some of the most loyal sports fans on Earth. They have gone through many seasons of victory, and more recently, long seasons of disappointment. 

But the 2021 season left many fans asking the same questions; are we really that bad? How are we losing with the talent we have? When taking a deeper look into the tragedy of the 3-9 2021 season, it is evident the Huskers should have pulled off more than a lousy three wins.  
```{r message=FALSE, warning=FALSE, echo=FALSE}
offensiveyards <- read_csv("offensiveyards.csv") %>% select(Name, `Pass Yards`) %>% rename(net_passing_yards = `Pass Yards`)

defensiveyards <- read_csv("defensiveyards.csv") %>% select(Name, `Pass Yards`) %>% rename(net_passing_yards_allowed = `Pass Yards`)

allpassyards <- offensiveyards %>% inner_join(defensiveyards)

threewins <- c("Nebraska", "Northwestern", "New Mexico", "Southern Mississippi", "Colorado State", "Louisiana Tech", "Duke", "Temple", "Georgia Tech", "Stanford", "Georgia Southern", "Ohio")

threewinpass <- allpassyards %>% filter(Name %in% threewins)

drives <- cfbd_drives(2021, team = "Nebraska", defense_conference = "B1G")

drivesNW <- cfbd_drives(2021, team = "Northwestern", defense_conference = "B1G")
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
plays <- cfbd_plays(2021, team = "Nebraska", defense_conference = "B1G")

NU <- cfbfastR::cfbd_game_team_stats(2021, team = "Nebraska")

NM <- cfbfastR::cfbd_game_team_stats(2021, team = "New Mexico")

NW <- cfbfastR::cfbd_game_team_stats(2021, team = "Northwestern")

CS <- cfbfastR::cfbd_game_team_stats(2021, team = "Colorado State")

LT <- cfbfastR::cfbd_game_team_stats(2021, team = "Louisiana Tech")

SM <-cfbfastR::cfbd_game_team_stats(2021, team = "Southern Mississippi")

DK <- cfbfastR::cfbd_game_team_stats(2021, team = "Duke")

TM <- cfbfastR::cfbd_game_team_stats(2021, team = "Temple")

GT <- cfbfastR::cfbd_game_team_stats(2021, team = "Georgia Tech")

SF <- cfbfastR::cfbd_game_team_stats(2021, team = "Stanford")

GS <- cfbfastR::cfbd_game_team_stats(2021, team = "Georgia Southern")

OH <- cfbfastR::cfbd_game_team_stats(2021, team = "Ohio")

games <- bind_rows(CS, LT, SM, NW, NM, NU, DK, TM, GT,SF, GS, OH)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
stats <- games %>%
  select(game_id, school, conference, points, total_yards, passing_tds, sacks, third_down_eff, third_down_eff_allowed,tackles, points_allowed, net_passing_yards, net_passing_yards_allowed, tackles, passing_tds) %>%
  separate(third_down_eff, into=c("third_down_converted","third_down_faced"), sep="-") %>% 
  separate(third_down_eff_allowed, into=c("third_down_converted_allowed","third_down_faced_allowed"), sep="-") %>%
  mutate_at(5:11, as.numeric) %>%
  mutate(
    net_passing_yards = as.numeric(net_passing_yards),
    net_passing_yards_allowed = as.numeric(net_passing_yards_allowed)
    )
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
allowed <- stats %>% select(game_id, school, conference, points_allowed)

defense <- stats %>% select(game_id, school, conference, sacks,tackles )

passing <- stats %>%  select(game_id, school, conference, net_passing_yards, net_passing_yards_allowed, passing_tds )

passing1 <- stats %>% 
  group_by(school) %>% 
  summarise(
    net_passing_yards = sum(net_passing_yards), 
    net_passing_yards_allowed = sum(net_passing_yards_allowed), 
    passing_tds = sum(passing_tds)
    ) %>% na.omit()
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
stats1 <- stats %>% 
  group_by(school) %>% 
  summarise(
    yards = sum(total_yards), 
    points = sum(points)
    ) %>% na.omit()

NUtotal <- stats1 %>% filter(school == "Nebraska")
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + 
  geom_bar(data=stats1, aes(x=reorder(school, yards), weight=yards)) +
  geom_bar(data= NUtotal, aes(x=reorder(school, yards), weight=yards), fill="red") +
  scale_y_continuous(labels=comma) +
  coord_flip() +
labs(title="Nebraska Knows How To Advance On The Field", subtitle="The Huskers clearly gained more yardage than their competitors. ", y= "Yards", x = "School" , caption = "Source:cfbfastR | By:Maddi Anderson") +   theme_minimal()+
  theme( 
    legend.position = "none", 
    panel.grid.major = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face="bold" , size= 18)
  )
```
When comparing Nebraska to their 3-9 competitors, the Huskers took the lead on gaining yards.They were number one out of all 3-9 college teams this season, and gained more than twice as many yards as the bottom team; New Mexico. Nebraska also ranked higher than another B1G team; Northwestern. 

We see that Nebraska excelled against the losing teams, but what about the rest of the NCAA? How did Nebraska compare to teams with a winning record? 
```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + 
  geom_point(data=allpassyards, aes(x=net_passing_yards, y=net_passing_yards_allowed), color="grey") + 
  geom_point(data=threewinpass, aes(x=net_passing_yards, y=net_passing_yards_allowed), color="red") +
  geom_vline(xintercept = 2987.646) +
  geom_hline(yintercept = 2911.254) +
  geom_text(aes(x=1500 , y=4000 , label= "Bad Offense,Bad Defense "), size=5, color="red") +
  geom_text(aes(x=5000 , y=2000 , label= "Good Offense,Good Defense "), size=5, color="red")  + 
  geom_text_repel(
    data=threewinpass, 
    aes(x=net_passing_yards, y=net_passing_yards_allowed, label=Name), size= 3)+
  labs(title="Nebraska Among The Best In Passing", 
  subtitle="The Huskers remain in the best quadrent, even after being 3-9.   ", y= "Passing Yards Allowed", x = "Passing Yards" , caption = "Source:cfbfastR | By:Maddi Anderson") +   theme_minimal()+
  theme( 
    legend.position = "none", 
    panel.grid.major = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face="bold" , size= 18)
  )
```
Taking a look at the good offense, good defense quadrant, we see two red dots that resemble two of our 3-9 teams; Colorado State and Nebraska. This past season Nebraska was keeping up with some of the best teams in college football stats wise, proving how strong the 2021 football team truly was. 
```{r message=FALSE, warning=FALSE, echo=FALSE}
thirds <- stats %>%
  group_by(school) %>% 
  summarise(
    third_down_faced = sum(third_down_faced), 
    third_down_converted = sum(third_down_converted))
```


```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot() + 
  geom_dumbbell(
    data=thirds, 
    aes(y=reorder(school, third_down_faced), x=third_down_faced, xend=third_down_converted),
    size = 2,
    colour = "grey",
    colour_x = "black",
    colour_xend = "red") +
  labs(title="Nebraska Is The Most Efficient On Third Downs", 
  subtitle="The red dot is third downs converted, while the black is third downs faced. The shorter the line the better.   ", y= "School", x = "Third Down Efficiency" , caption = "Source:cfbfastR | By:Maddi Anderson") +   theme_minimal()+
  theme( 
    legend.position = "none", 
    panel.grid.major = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face="bold" , size= 18)
  )

```
When it comes to efficient third downs, Nebraska was, well, mediocre. Nebraska may have the shortest line out of the 3-9 teams, but these are some of the worst teams that played this season. If Nebraska wants to improve their record, they need to start improving their third down efficiency. 

After considering the statistics, Nebraska played better than a 3-9 team should. They dominated their 3-9 competitors and kept up with the big dogs in passing yards. The upcoming 2022 season will prove if this season was a lucky one for the Huskers, or if they are as talented as their numbers say.